\chapter{Approximation Theory}
The main purpose of Approximation theory is to replace a complicated function by one which is simpler and more
manageable.
\section{Lagrange}
Set of point $(x_i,y_i)$ all i $0\leq i \leq n $ i.e. n+1 points.
Define a polynomial of degree n, which satisfies all points.

Assume there exists a poly of degree n which has the properties

\begin{eqnarray}
L_{i}(x_j) = 0 \hspace{0.2 in}
L_{i}(x_i) = 1 
\label{Lagrange poly 1}
\end{eqnarray}
We may then write
\[p_n(x) = \sum_{i=0}^{n} L_{i}(x)y_{i} \]

Now to define $L_{i}$.  If $L_i$ is to satisfy criteria then it must contain a factor
\[ (x-x_0)(x-x_1)...(x-x_{i-1})(x-x_{i+1})...(x-x_n) \]
Since it has n terms it is a poly of degree n.  Therefore
\[ L_i = A_i(x-x_0)(x-x_1)...(x-x_{i-1})(x-x_{i+1})...(x-x_n) \]
$A_i $ is chosen such that (s.t.) (\ref{Lagrange poly 1}) is satisfied.
\[A_{i} = \frac{1}{(x-x_0)(x-x_1)...(x-x_{i-1})(x-x_{i+1})...(x-x_n)} \]
\[L_{i} = \Pi_{j=0,j \neq i}^{n} \frac{x-x_j}{x_i-x_j} \]

To prove the uniqueness suppose that there is another poly $q_n$ of degree at most n s.t. $q_{n}(x_{i}) = y_i$.\\
Now $p_n - q_n$ is itself a poly of degree at most n and vanishes at n+1 points $x_i$ because
\[ p_n(x_i) - q_n(x_i) = y_i - y_i =0 \]
However by the fundamental theorem of Algebra a poly of degree n has at most n roots unless it is identically zero. Hence $q_n = p_n$.
\subsection*{Example:}
\begin{quote}
\begin{tabular}{lrrr}
x & 0 & 1 & 3 \\ \hline
f(x) & 1 & 3 & 2 
\end{tabular}
\end{quote}
\[ L_0 = \frac{(x-1)(x-3)}{(0-1)(0-3)},
 L_1 = \frac{(x-0)(x-3)}{(1-0)(1-3)},
 L_2 = \frac{(x-0)(x-1)}{(3-0)(3-1)} \]

\[ P_2(2) = 1.L_0(2)+3.L_1(2)+2.L_2(2) = \frac{10}{3} \]

\begin{theorem}
\label{Lagrange Thm}
Let [a,b] be any interval containing the distinct points $x_i$ (i=0,1,..,n).  If all derivatives of f up to and including $f^{n+1}$ exist and are continuous on
[a,b] the corresponding to each value of x in [a,b] $\exists$ a number $\xi$
in (a,b) s.t.
\[ f(x) - p_n(x) = (x-x_0)(x-x_1)...(x-x_n) \frac{f^{n+1}(\xi)}{(n+1)!} \]
\end{theorem}
\begin{proof}
Exercise
\end{proof}
\textbf{Similar} \textbf{interpolation:}
Divided Difference,
Finite Difference form,
Hermite Interpolation.
\section{Splines}
\begin{definition} 
\emph{Spline function}
Given a strictly increasing sequence of real numbers $a=x_0,x_1,...,x_n=n$ a 
spline function S(x) of degree m with the knots $x_0,...x_n$ is a function of order $m \geq 1 $ defined on the entire real line having the following two properties:
\begin{enumerate}
\item
S(x) is a polynomial of degree $< m$ on each subinterval $[x_{i-1},x_i]$;
\item
$S^{r}(x)$ is continuous on $[a,b]$ for $0\leq r \leq m-1$.
\end{enumerate} 
\end{definition}

\underline{Cubic Spline} is the most popular for m=4.
From this point on we only deal with Cubic Splines.
\begin{equation}
S_i(x) = a_i(x-x_i)^3 + b_i(x-x_i)^2+c_i(x-x_i)+d_i
\label{S_i original}
\end{equation}
\begin{subequations}
\label{S_i conditions}
\begin{equation}
\label{S_i conditions a}
S_i(x_i) = y_i \ \ i=0,1,..,n-1 \ S_{n-1}(x_n) = y_n;
\end{equation}
\begin{equation}
\label{S_i conditions b}
S_i(x_{i+1}) = S_{i+1}(x_{i+1})  i=0,1,..,n-2
\end{equation}
\begin{equation}
\label{S_i conditions c}
S_i^{'}(x_{i+1}) = S_{i+1}^{'}(x_{i+1}) i=0,1,..,n-2
\end{equation}
\begin{equation}
\label{S_i conditions d}
S_i^{''}(x_{i+1}) = S_{i+1}^{''}(x_{i+1}) i=0,1,..,n-2
\end{equation}
\end{subequations}
If there are n+1 points the number of intervals  and number of equations are n.
There are 4n unknowns which are the $a_i,b_i,c_i,d_i$ for i=0,1,...,n-1.
Immediately \ref{S_i conditions a}
gives
\begin{equation}
\label{d_i}
d_i = y_i
\end{equation}

(\ref{S_i conditions b}) gives
\begin{eqnarray*}
y_{i+1} &= &a_i(x_{i+1}-x_i)^3 + b_i(x_{i+1}-x_i)^2+c_i(x_{i+1}-x_i)+y_i \\
& = &  a_ih_i^3 + b_ih_i^2+c_ih_i+y_i \ \ i=0,...,n-1\\
h_i=(x_{i+1}-x_i)
\end{eqnarray*}
Differentiate (\ref{S_i original})

\begin{equation}
\label{S_i derivatives}
S_i^{'}(x) = 3a_i(x-x_i)^2 + 2b_i(x-x_i)+c_i\\
S_i^{''}(x) = 6a_i(x-x_i) + 2b_i \ \ for \ \ i=0,...,n-1\\
\end{equation}
Development is simplified if we write in terms of the 2nd derivative that is, if we let $M_i = S^{''}$ for i=0,..n-1 and $M_n = S_{n-1}^{''}(x_{n})$

\begin{eqnarray*}
M_i &=& 6a_i(x_i-x_i) + 2b_i\\
& = & 2 b_i \\
M_{i+1} &=& 6a_i(x_{i+1}-x_i) + 2b_i\\
& = &6a_ih_i+ 2 b_i
\end{eqnarray*}
Hence we write
\begin{equation}
\label{b_i}
b_i = \frac{M_i}{2}
\end{equation}
\begin{equation}
\label{a_i}
a_i = \frac{M_{i+1}- M_i}{6h_i}
\end{equation}
We substitute the relations for $a_i,b_i,c_i,d_i$ given (\ref{a_i}),(\ref{b_i}) and
(\ref{d_i}) into (\ref{S_i original}) to get.
\begin{eqnarray*}
y_{i+1} & = & \frac{(M_{i+1}-M_{i})}{6h_i}h^3_i + \frac{M_i}{2}h_i^2+c_ih_i+y_i \\
c_i & = & \frac{y_{i+1}-y_{i}}{h_i} - \frac{2h_iM_i+h_iM_{i+1}}{6} 
\end{eqnarray*}
We now evoke the condition that the slopes of the 2 cubics that join at the point $(x_i,y_i)$ are the same (\ref{S_i conditions c}).
\begin{equation*}
y_i^{'} = 3a_i(x_i-x_i)^2 + 2b_i(x_i-x_i)+c_i = c_i
\end{equation*}
In the previous interval from $x_{i-1}$ to $x_i$ the slope at its right end will
be 
\begin{equation*}
y_i^{'} = 3a_{i-1}h_{i-1}^2 + 2b_{i-1}h_{i-1}+c_{i-1} 
\end{equation*}
Equating these and replacing a,b,c,d we get

\begin{eqnarray}
\label{M form}
h_{i-1}M_{i-1} + 2(h_{i-1}+h_{i})M_i+h_iM_{i+1} &= &6(\frac{y_{i+1} - y_{i}}{h_i} -\frac{y_{i}-y_{i-1}}{h_{i-1}})\\
& = & 6(f[x_i,x_{i+1}] - f[x_{i-1},x_{i}])
\end{eqnarray}
This gives us n-1 equations and n+1 unknowns.
Writing this in Matrix form:
\begin{tiny}
\[ 
	\left( \begin{array}{ccccccc}
    h_0 & 2(h_0+h_1) & h_1 & & & &\\
     & h_1 & 2(h_1+h_2) & h_2& & &\\
     &  & h_2 & 2(h_2+h_3)&h_3 & &\\
     &  & . & & & &\\
     &  &  &. & & &\\
     &  &  & &. & &\\
     &  &  & h_{n-2} &2(h_{n-2}-h_{n-1}) &h_{n-1} &\end{array} \right)
	\left( \begin{array}{c}
      M_0\\
      M_1\\
      M_2\\
      M_3\\
      .\\
      .\\
      .\\
M_{n-1}	\\
      M_n\end{array} \right)
=
6	\left( \begin{array}{c}
      f[x_1,x_2]-f[x_0,x_1]\\
    f[x_2,x_3]-f[x_1,x_2]  \\
   f[x_3,x_4]-f[x_2,x_3]   \\
      .\\
      .\\
      .\\
  f[x_{n-1},x_{n}]-f[x_{n-2},x_{n-1}]    \end{array} \right)
\] 
\end{tiny}
We need 2 more equations to solve to do this we introduce conditions:
\begin{enumerate}
\item
Take $M_0=0$ and $M_n=0$.  This makes the end cubics approach linearity at the extremities.  This condition, called a natural spline, matches precisely to the 
drafting device. This technique is used very frequently.
\[ 
	\left( \begin{array}{cccccc}
      2(h_0+h_1) & h_1 & & & &\\
      h_1 & 2(h_1+h_2) & h_2& & &\\
       & h_2 & 2(h_2+h_3)&h_3 & &\\
       & . & & & &\\
       &  &. & & &\\
       &  & &. & &\\
      & &  & h_{n-2} &2(h_{n-2}-h_{n-1}) &\end{array} \right)
\]

\item
Another often used condition is to force the slopes at each end to assume specified values.  When that is not known, the slope might be estimated from the points.
If $f^{'}(x_0)=A$ and $f^{'}(x_n)=B$, we use these relations:

\begin{eqnarray}\label{end points} \mbox{At the Left end:  }& 2h_0M_0+h_1M_1=6(f[x_0,x_1]-A). \\
\mbox{At the right end:  } &h_{n-1}M_{n-1}+2h_{n}M_{n}=6(B-f[x_{n-1},x_n]). \end{eqnarray}

\[ 
	\left( \begin{array}{cccccc}
      2(h_0) & h_1 & & & &\\
      h_1 & 2(h_1+h_2) & h_2& & &\\
       & h_2 & 2(h_2+h_3)&h_3 & &\\
       & . & & & &\\
       &  &. & & &\\
       &  & &. & &\\
      & &  & h_{n-2} &2(h_{n-1}) &\end{array} \right)
\]

\item
Take $S_0=S_1$, $S_{n-1}=S_{n-2}$.  This is equivalent  to assuming that the end cubics approach parabolas at their extremities
\[ 
	\left( \begin{array}{cccccc}
      (3h_0+2h_1) & h_1 & & & &\\
      h_1 & 2(h_1+h_2) & h_2& & &\\
       & h_2 & 2(h_2+h_3)&h_3 & &\\
       & . & & & &\\
       &  &. & & &\\
       &  & &. & &\\
      & &  & h_{n-2} &(2h_{n-2}-3h_{n-1}) &\end{array} \right)
\]

\end{enumerate}
Rephrasing equation (\ref{M form}) such that we have:
\[\mu_jM_{j-1} +2M_j +\lambda_{j+1}M_{j+1} = d_j \]
\[ \lambda_j = \frac{h_{j+1}}{h_{j}+h_{j+1}} \, \mu_1=1-\lambda_j=\frac{h_j}{h_j+h_{j+1}} \]
\[d_j = \frac{6}{h_j+h_{j+1}}\left(f[x_{j},x_{j+1}]-f[x_{j-1},x_{j}]\right) \ \
j=1,...,n-1 \]
\begin{theorem}
If $f\in C^{4}[a,b]$ and $|f^{4}(x)|\leq L$ for $x \in [a,b]$ then,
\[
||M-F||\leq ||r|| \leq \frac{3}{4}L||\Delta||^{2} \]
\end{theorem}
Notation:
\[
M=\left( \begin{array}{c}
      M_0\\
      M_1\\
      .\\
      .\\
      .\\
      M_n\end{array} \right)
F=
\left( \begin{array}{c}
      f^{''}(x_0)\\
      f^{''}(x_1)\\
      .\\
      .\\
      .\\
      f^{''}(x_n)\end{array} \right)
\]
\[r=d-AF \]
\[||\Delta||=max_j|x_{j+1}-x_{j}| \]

\begin{proof}
By defn $r_0=d_0-2f^{''}(x_0)-f^{''}(x_1)$ and from (\ref{end points})
\[
r_0 = \frac{6}{h_1}\left(\frac{y_1-y_0}{h_1} - y'_0\right) -2f^{''}(x_0) - f^{''}(x_1) \]
Using Taylor expansion $y_1=f(x_1)$ and $f^{''}(x)$ in terms of the value and the derivatives of the function f at $x_0$ yields.
\begin{eqnarray*}
r_0&=& \frac{6}{h_1}(f^{'}(x_0)+\frac{h_{1}}{2}f^{''}(x_0)+\frac{h_1^2}{6}f^{'''}(x_0) + \frac{h_{1}^3}{24}f^{4}(\tau_1)-f^{'}(x_0) )\\
& & -2f^{''}(x_0) - [f^{''}x_0+h_1f^{''}(x_0)+\frac{h_1^2}{2}f^{4}(\tau_2)]\\
&=& \frac{h_1^2}{4}f^{4}(\tau_1) - \frac{h_{1}^2}{2}f^{4}(\tau_2)
\end{eqnarray*}
with $\tau_1,\tau_2 \in [x_0,x_1]$. Hence
\[ |r_0| \leq \frac{3}{4} L ||\Delta||^2 \]
Analogously we find
\[
r_n = d_n - f^{''}(x_{n-1})-2f^{''}(x_n) \]
that 
\[ |r_n| \leq \frac{3}{4} L ||\Delta||^2 \]
For the remaining components of $r=d-AF$ we find similarly
\begin{eqnarray*}
r_j&=&d_j - \mu_jf^{''}(x_{j-1})-2f^{''}(x_j) -\lambda_{j+1}f^{''}(x_{j+1})\\
&=&\frac{6}{h_j+h_{j+1}}\left(f[x_{j},x_{j+1}]-f[x_{j-1},x_{j}]\right)\\ & & -
\frac{h_{j}}{h_{j}+h_{j+1}}f^{''}(x_{j-1}) - 2f^{''}(x_j)
\frac{h_{j+1}}{h_{j}+h_{j+1}}f^{''}(x_{j+1}) 
\end{eqnarray*}
Taylors formula at $x_j$ then gives
\begin{eqnarray*}
r_j & =& \frac{1.0}{h_j+h_{j+1}}\left[ \frac{h_{j+1}^3}{4}f^{4}(\tau_1 )  \frac{h_{j+1}^3}{4}f^{4}(\tau_2 )  -\frac{h_{j+1}^3}{4}f^{4}(\tau_3 )  -\frac{h_{j+1}^3}{4}f^{4}(\tau_4 ) \right]
\end{eqnarray*}
where $\tau_1, \tau_2, \tau_3,\tau_4 \in [x_{j-1},x_{j+1}],$
\[ |r_j| \leq \frac{3}{4}L \frac{[h_{j+1}^3+h_j^3]}{h_j+h_{j+1}}\leq \frac{3}{4}L||\Delta||^2 \]
for j=1,...,n-1
\[
||r|| \leq \frac{3}{4} L || \Delta ||^2 \]
and since $r=A(M-F)$ implies $||M-F|| \leq ||r||$ using the non-singular properties of A.
\end{proof} $\bigodot$
\begin{theorem}
Suppose 
$f \in C^{4}[a,b]$ and $ |f^{4}(x)| \leq L $for $x \in [a,b]$. Let $\Delta$ be a partition
$\Delta = \{ a=x_0 < ... < x_n = b \} $ of the interval $[a,b]$, and K be a constant
\[
\frac{||\Delta||}{|x_{j+1}-x_{j}|} \leq K
\]
If $S_{\Delta}$ is the Spline function which interpolates the values of the function f at the knots $x_0,...x_n \in \Delta$
and satisfies $S^{'}_{\Delta} = f^{'}(x)$ for $(a,b) $ then there exists constants $C_{k} \leq 2$ which do depend on the partition $\Delta$ such that for $ x \in [a,b]$
\[
|f^{(k)}(x)-S_{\Delta}^{k}(x)|\leq C_{k}LK||\Delta||^{4-k} \]
for k=0,1,2,3.
\end{theorem}
\begin{proof}
We prove the proposition first for k=3.
For $x \in [x_{j-1},x_j]$
\begin{eqnarray*}
S^{'''}_{\Delta}(x)-f^{'''}(x)& = &\frac{M_j-M_{j-1}}{h_j} - f^{'''}(x)\\
& = & \frac{M_j - f^{''}(x_j)}{h_j} -\frac{M_{j-1}-f^{''}(x_{j-1})}{h_j}\\
& & + \frac{f^{''}(x_j)-f^{''}(x) - [f^{''}(x_{j-1}-f^{''}(x)]}{h_j} -f^{'''}(x)
\end{eqnarray*}
Using previous proof and Taylors Thm at we conclude
\begin{eqnarray*}
 ||S^{'''}_{\Delta}(x)-f^{'''}(x)|| &\leq& \frac{3}{2}L\frac{||\Delta||^2}{h_j}\\
& & + \frac{1}{h_j} \left|(x_j-x)f^{'''}(x) + \frac{(x_j-x)^2}{2} f^4(\eta_1)\right.\\
& & \left.- (x_{j-1}-x)f^{'''}(x) + \frac{(x_{j-1}-x)^2}{2} f^4(\eta_2) - h_jf^{'''}(x) \right| \\
&\leq& \frac{3}{2} L \frac{||\Delta||^2}{h_j} + \frac{L}{2} \frac{||\Delta||^{2}}{h_j} \ \ \ \eta_1,\eta_2 \in [x_{j-1},x_{j}]
\end{eqnarray*}
By hypothesis $\frac{||\Delta||}{h_j} \leq K $ for every interval.
Thus $|f^{'''}(x) - S^{'''}_{\Delta}(x)| \leq 2LK||\Delta||$
To prove the proposition for k=2 we observe for each $x \in (a,b)$ a closest
knot $x_j = x_j(x)$ for which $|x_j(x) - x| \leq \frac{1}{2}||\Delta||$.
\begin{eqnarray*}
f^{''}(x) - S_{\Delta}^{''}(x) & = & f^{''}(x_j(x)) - S_{\Delta}^{''}(x_j(x)) 
+ \int^{x}_{x(x_j)}(f^{'''}(t)-S_{\Delta}^{'''}(t))dt
\end{eqnarray*}
and since $K \geq 1$

\begin{eqnarray*}
|f^{''}(x) - S_{\Delta}^{''}(x)| & \leq & \frac{3}{4} L ||\Delta||^2 +\frac{1}{2}||\Delta||2LK||\Delta|| \\
& \leq & \frac{7}{4}LK||\Delta||^2  \ \ x \in [a,b]
\end{eqnarray*}
We consider k=1 next. In addition to the boundary points $\xi_0 = a, \xi_{n+1}=b$
$\exists$ by Rolle's Theorem, n further points $\xi \in (x_{j-1},x_j),\ \ j=1,...,n$
\[
f^{'}(\xi_j) = S_{\Delta}^{'}(\xi_j) \ \ j=0,..,n+1
\]
For $x \in [a,b] \ \exists$ a closest of the above points $\xi_j = \xi_j(x)$, for
which consequently 
\[
|\xi_j(x)-x| \leq ||\Delta||
\]

Thus
\[
f^{'}(x)-S^{'}_{\Delta}(x) = \int_{\xi_j(x)}^{x}(f^{''}(t)-S^{''}_{\Delta}(t))dt
\]
\begin{eqnarray*}
|f^{'}(x) - S_{\Delta}^{'}(x)| & \leq & \frac{7}{4} L K ||\Delta||^2 ||\Delta||\\
& \leq & \frac{7}{4}LK||\Delta||^3  \ \ x \in [a,b]
\end{eqnarray*}
The case k=0. Since 
\[
f(x)-S_{\Delta}(x) = \int_{\xi_j(x)}^{x}(f^{'}(t)-S^{'}_{\Delta}(t))dt
\]
It follows from the above result for k=1 that 
\begin{eqnarray*}
|f(x) - S_{\Delta}(x)| & \leq & \frac{7}{4} L K ||\Delta||^3 \frac{1}{2}||\Delta||\\
& \leq & \frac{7}{8}LK||\Delta||^4  \ \ x \in [a,b]
\end{eqnarray*}
\end{proof} $\bigodot$
\section{Bezier Curves and B-Splines}
Bezier curves named after the French engineer P. Bezier of Renault.
For the curves the points are more like "control" points, we select them to determine
the shape of the curve we are working on.\\
We consider mainly the cubic version of these curves. We will express $y=f(x)$ in parametric form.
The parametric form represents a relation between x and y by two equations
$x=F_1(u), \ y=F_2(u)$, u is called the parameter.\\
When x and y are expressed in terms of a parameter $u \  (x(u),y(u)) \ 0\leq u \leq 1$
defines a set of points (x,y) associated with the values of u.\\
Control points $P_i = (x_i,y_i) \ i=0,1,..,n$.\\
The nth degree Bezier polynomial determined by n+1 control points is given by
\[
P(u) = \sum^{n}_{i=0}
\left( \begin{array}{c}
      n \\
      i\end{array} \right)
(1-u)^{n-i}u^i P_i \]
\[
\left( \begin{array}{c}
      n \\
      i\end{array} \right)
= \frac{n!}{i!(n-i)!}
 \]
For n=3,
\[
x(u) = (1-u)^3x_0+3(1-u)^2ux_1+3(1-u)u^2x_2+u^3x_3
\]
\[
y(u) = (1-u)^3y_0+3(1-u)^2uy_1+3(1-u)u^2y_2+u^3y_3
\]
\textbf{Properties of a Bezier Curve}
\begin{enumerate}
\item
$P(0)=P_0$ \hspace{0.2 in} $P(1)=P_3$;
\item
Since $\frac{dx}{du}=3(x_1-x_0)$
and $\frac{dy}{du}=3(y_1-y_0)$ at u=0, the slope of the curve at u=0 is 
$\frac{dy}{dx}=\frac{y_1-y_0}{x_1-x_0}$ which  is the slope at the secant line 
between $P_0$ and $P_1$.  Similarly, for the slope at u=1.
\item
The curve is contained in the convex hull determined by the 4 points.
\end{enumerate}
\subsection*{B-Splines}
$P_i = (x_i,y_i)$ i=0,..,n  cubic B-splines for the interval $P_i,P_{i+1}$ i =1,..n-1 is
\begin{eqnarray*}
B_i(u) & = & \sum^{2}_{k=-1} b_k P_{i+k} \\
b_{-1} & = & \frac{(1-u)^3}{6} \\
b_{0} & = & \frac{u^3}{2}-u^2+\frac{2}{3} \\
b_{1} & = & \frac{-u^3}{2}-\frac{u^2}{2}+\frac{u}{2}+\frac{1}{6} \\
b_{2} & = & \frac{u^3}{6} \ \ 0 \leq u \leq 1 \\
\end{eqnarray*}
\begin{equation*}
B_i(u)  = 
 (\frac{(1-u)^3}{6}) P_{i-1} +
 (\frac{u^3}{2}-u^2+\frac{2}{3}) P_{i} +
 (\frac{-u^3}{2}-\frac{u^2}{2}+\frac{u}{2}+\frac{1}{6}) P_{i_1}
 +(\frac{u^3}{6})P_{i+2}
\end{equation*}
\textbf{Properties of a B-Spline}
\begin{enumerate}
\item
B-Splines are pieced together so they agree at their joints in three ways
\begin{eqnarray*}
B_i(1)=B_{i+1}(0)&=&\frac{P_i+4P_{i+1}+P_{i+2}}{6}\\
B_i^{'}(1)=B_{i+1}^{'}(0)&=&\frac{-P_i+P_{i+2}}{2}\\
B_i^{''}(1)=B_{i+1}^{''}(0)&=& P_i-2P_{i+1}+P_{i+2}
\end{eqnarray*}
\item
The portion of the curve determined by each group of four points is within
the convex hull of these points.
\end{enumerate}
\textbf{End Points}
While our system of polynomials are defined for i=1 to i=n-1 we need to 
create extra points s.t. the end points are include.\\
We create $P_{-2},P_{-1},P_{n+1},P_{n+2}$s.t. $P_0 = P_{-1}=P_{-2}$ and
$P_{n} = P_{n+1} = P_{n+2}$.\\
\textbf{Matrix Form:}
\begin{eqnarray*}
B_{i}(u) &=& \frac{1}{6}[u^3,u^2,u^1,1]
\left[\begin{array}{cccc}
-1 & 3 & -3 & 1 \\
3 & -6 & 3 & 0 \\
-3 & 0 & 3 & 0 \\
1 & 4 & 1 & 1 \end{array} \right]
\left[\begin{array}{c}
P_{i-1} \\
P_{i} \\
P_{i+1} \\
P_{i+2} \end{array} \right]\\
&=&\frac{u^{T}M_b P}{6}
\end{eqnarray*}
\section{Linear square approximation}
There is no justification for insisting that the data points be reproduced exactly, and such and approximation may well be very misleading since unwanted oscillations are likely.  A more satisfactory approach would be to find a line which
passes closest to all points.
Formula for the line.
\[y_{i} = ax_i +b \]
Let $e_i$ denote the error in relation to each point.
\begin{eqnarray*}
 S &= &e_{1}^2 + e_{2}^2+ . . . + e_{n}^2 \\
& = & \sum_{i=1}^{n}e_i^2 \\
& = & \sum_{i=1}^{n}(y_i-ax_i-b)^2
\end{eqnarray*}
To minimise the error we calculate a and b by differentiating S wrt to a and b
respectively and setting to 0.
\begin{eqnarray*}
\frac{\delta S}{\delta a}& = 0 =& \sum_{i=1}^{n}2(y_i-ax_i-b)(-x_i)\\
\frac{\delta S}{\delta b}& = 0 =& \sum_{i=1}^{n}2(y_i-ax_i-b)(-1)
\end{eqnarray*}
Rearranging these equations we have 2 equations a 2 unknowns of the form.
\begin{eqnarray*}
a\sum x_{i}^2 + b \sum x_i &= & \sum x_i y_i \\
a\sum x_{i}^2 + b N &= & \sum  y_i 
\end{eqnarray*}
This can be solved easily.  Now for a more general form.
\begin{eqnarray*}
y & = & a_0 +a_1 x + ... + a_n x^n \\
e_i & = Y_i-y_i= & Y_i - a_0 -a_1 x_i - ... - a_n x_i^n \\
S = & \sum_{i=1}^n e_i^2 = &\sum_{i=1}^n (Y_i - a_0 -a_1 x_i - ... - a_n x_i^n)^2 \\
\end{eqnarray*}
Differentiating w.r.t. each of the constants $a_i$ for $i=0,..,n$ and setting each of these to minimise error.
\begin{eqnarray*}
\frac{\delta S}{\delta a_0}&= 0 =& \sum_{i=1}^{n}2(Y_i - a_0 -a_1 x_i - ... - a_n x_i^n)(-1) \\
\frac{\delta S}{\delta a_1}&= 0 =& \sum_{i=1}^{n}2(Y_i - a_0 -a_1 x_i - ... - a_n x_i^n) (-x_i)\\
.&.&.\\
.&.&.\\
.&.&.\\
\frac{\delta S}{\delta a_n}&= 0 =& \sum_{i=1}^{n}2(Y_i - a_0 -a_1 x_i - ... - a_n x_i^n) (-x_i^n)
\end{eqnarray*}
putting this into matrix form.
\begin{equation}\left[
\begin{array}{ccccccc}
N&\sum x_i & \sum x_i^2 & \sum x_i^3 &.&.& \sum x^n_i\\
\sum x_i & \sum x_i^2 & \sum x_i^3 &.&.& .&\sum x^{n+1}_i\\
.&.& .&.& .&.& .\\
.&.& .&.& .&.& .\\
\sum x_i^n & \sum x_i^{n+1} & \sum x_i^{n+2} &.&.& .&\sum x^{2n}_i
\end{array}\right]
a = 
\left[
\begin{array}{c}\sum Y_i\\
.\\
.\\
.\\
\sum x_i^nY_i\end{array}\right]
\end{equation}
This matrix is ill conditioned at n=4,5 there is no real problem but as n gets larger
round-off error will arise.
\section{Trigonometric Interpolation}
Trigonometric interpolation is for periodic functions, $f(t+2\pi)=f(t)$.
Using combinations of cos(hx) and sin(hx) for integer h.
\begin{equation}
\label{Trig inter}
\phi(x) = \frac{A_0}{2} + \sum_{h=1}^M(A_{h}cos(hx) + B_h sin(hx))
\end{equation}
\begin{equation}
\phi(x) = \frac{A_0}{2} + \sum_{h=1}^{M-1}(A_{h}cos(hx) + B_h sin(hx)) +\frac{A_M}{2}cos(Mx)
\end{equation}
respectively N = 2M+1 (odd) and N=2M.\\
As before we require
\[P_{n}(t_i)=f(t_i) \ \ \ i=0,..N \]
But due to the periodic nature of f we require the nodes to lie
$0\leq t \leq \pi \ \ (or \ -\pi \leq t \leq \pi)$
\[0=t_0 < t_1 < ... < t_n = 2 \pi \]
\[x_k = \frac{2\pi k}{N} \ \ k=0,..,N-1 \]
De Moivre' formula:
\[
e^{i\theta}=cos\theta+isin\theta \ \ i-imaginary
\]
\[cos\theta =\frac{ e^{i\theta}  +e^{-i\theta}}{2} \ \ sin\theta = \frac{ e^{i\theta}  -e^{-i\theta}}{2}
\]
\begin{equation}
p(x) = \beta_0+\beta_1 e^{ix}+...+\beta_{N-1}e^{N-1}ix
\end{equation}
with complex coefficients $\beta_j$ s.t.
\[p(x_k)=f_k \ \ k=0,...,N-1 \]
Indeed 
\[
e^{-hix_k}=
e^{-2\pi ihk/N}=
e^{(N-h)ix_k}
\]
\[
cos(hx_k) = \frac{ e^{hix_k}  +e^{(N-h)ix_k}}{2}
\ \ sin(hx_k) = \frac{ e^{hix_k}  -e^{(N-h)ix_k}}{2}
\]
Making these substitutions into (\ref{Trig inter}).\\
We can express $A_h,B_h$ in terms of $\beta_j$.
\begin{enumerate}
\item
If N is odd then N=2M+1
\[\beta_0=\frac{A_0}{2} \ \ \beta_j=\frac{1}{2}(A_j-iB_j), \ \ \beta_{N-j}=\frac{1}{2}(A_j+iB_j) \ \ j=1,..,M
\]
\[A_0=2\beta_0 \ \  A_h=\beta_h+\beta_{N-h}, \ \ B_{h}=i(\beta_h+\beta{N-h}) \ \ j=1,..,M
\]
\item
If N is even N=2M
\[\beta_0=\frac{A_0}{2} \ \ \beta_j=\frac{1}{2}(A_j-iB_j), \ \ \beta_{N-j}=\frac{1}{2}(A_j+iB_j) \ \ j=1,..,M-1 \beta_{M}=\frac{A_{M}}{2}
\]
\[A_0=2\beta_0 \ \ A_h=\beta_h+\beta_{N-h}, \ \ B_{h}=i(\beta_h+\beta{N-h}) \ \ j=1,..,M
\]
\end{enumerate}
$\phi(x)$ and its corresponding exponential poly p(x) cores pond for all support arguments $x_k=2\pi k N$ of equidistant partition of the interval $[0,2\pi]$
\[
f_k=\phi(x_k)=p=(x_k) \ \ \ k =0,...,N-1
\]
However $\phi(x)=p(x)$ need not hold for intermediate points $x\not=x_k$.  The
two interpolant problems are equivalent in so far as a solution to one problem will
produce a solution to the other via the coefficients.\\
p(x) has a simpler structure than $\phi(x)$.
Abbreviating
\[ \omega = e^{ix} \ \ \omega_k=e^{ix_k}\]
\[P(\omega) = \beta_0+\beta_i \omega +...+\beta_{N-1}\omega^{N-1} \]
and since $\omega_j \not= \omega_k \ \ 0\leq j \leq k\leq N-1$ it becomes clear we 
are just faced with a standard poly interpolation problem in disguise.\\
\begin{theorem}
For any support points $(x_{k},f_{k})$ k=0,,..N-1 with $f_k$ complex and $x_k = 2\pi k\backslash N$ there
exists a unique phase polynomial
\[ p(x) = \beta_0 + \beta_1 e^{ix} +...+ \beta_{N-1} e^{(N-1)ix} \]
with
\[ p(x_k)=f_k \]
for k=0,1,..,N-1\\
\end{theorem}
\begin{theorem}
If the integer r is not a multiple of 2m then
\[
\sum_{j=0}^{2m-1} cosrx_j =0 \ and \ 
\sum_{j=0}^{2m-1} sinrx_j =0 
\]
Also
\[
\sum_{j=0}^{2m-1} (cosrx_j)^2 =m \ and \ 
\sum_{j=0}^{2m-1} (sinrx_j)^2 =m 
\]
\\
\end{theorem}
\begin{proof}
\[
\sum_{j=0}^{2m-1} (cosrx_j) +
i\sum_{j=0}^{2m-1} (sinrx_j) =
\sum_{j=0}^{2m-1} [(cosrx_j) +i(sinrx_j)] =
\sum_{j=0}^{2m-1}e^{irx_j} 
\]
But $e^{irx_j}=e^{ir2\pi j/N}$. Since 
$\sum_{j=0}^{2m-1}e^{irx_j} $ is a geometric series with first term 1 and ratio
$e^{ir2\pi/N}\not=1$ we have 
\[ \sum_{j=0}^{2m-1}e^{irx_j} =\frac{1-(e^{ir\pi/M})^{N}}{1-e^{ir2\pi/N}}
 =\frac{1-(e^{2ir\pi})}{1-e^{ir2\pi/N}}
 \]
But $e^{ir2\pi}=cos(2r\pi)+sin(2r\pi)=1$, so
\[
\sum_{j=0}^{2m-1} (cosrx_j) +
i\sum_{j=0}^{2m-1} (sinrx_j) = 0 \]
which implies
\[
\sum_{j=0}^{2m-1} (cosrx_j) =0
\sum_{j=0}^{2m-1} (sinrx_j) = 0 \]

\[
\sum_{j=0}^{2m-1} (cosrx_j)^2 =
\sum_{j=0}^{2m-1} \frac{1}{2}(1+cos2rx_j) =\frac{1}{2}(2m+0)=m
\]
Also
\[
\sum_{j=0}^{2m-1} (sinrx_j)^2 =m
\]
\end{proof}$\bigodot$
\begin{theorem}
The phase polynomial $p(x) = \sum_{j=0}^{N-1} \beta_j e^{jix}$ satisfies 
\[ p(x_k)=f_k \]
for $f_k$ complex and $x_k = 2\pi k \backslash N$ iff
\[
\beta_j = \frac{1}{N} \sum_{k=0}^{N-1} f_k \omega^{-j}_k = \frac{1}{N} \sum_{k=0}^{N-1} f_k e^{-2 \pi jk\backslash N}
\]
j=0,1,.. N-1.
\\
\end{theorem}

\begin{proof}
Using the vector notation
\[f=(f_0,f_1,...f_{N-1})^{T}\]
\[\frac{1}{N}\sum_{k=0}^{N-1}f_{k}\omega^{-j}_k=\frac{1}{N}[f,\omega^j]
=\frac{1}{N}[\beta_0\omega+...+\beta_{N-1}\omega^{N-1},\omega^{j}]=\beta_j
\]
\end{proof}$\bigodot$

\begin{theorem}
The trigonometric expressions 
\[
\phi(x) = \frac{A_0}{2} + \sum^{m}_{h=0} (A_h cos hx + B_h sin hx) \]
\[
\phi(x) = \frac{A_0}{2} + \sum^{m-1}_{h=0} (A_h cos hx + B_h sin hx)+\frac{A_{m}}{2}cosmx \]

where N=2m+1 and N=2m respectively satisfy
\[\phi(x_k) = f_k \ \ k=0,1,..,N-1 \]
for $x_k=2 \pi k \backslash N $ iff the coefficients of $\phi(x)$ are given by
\[ A_h = \frac{2}{N} \sum_{k=0}^{N-1} f_k cos x_k h = \frac{2}{N} \sum_{k=0}^{N-1} f_k cos 2\pi k \backslash N h \]
\[
B_h = \frac{2}{N} \sum_{k=0}^{N-1} f_k sin x_k h = \frac{2}{N} \sum_{k=0}^{N-1} f_k sin 2\pi k \backslash N h \]
\end{theorem}
\begin{proof}
\[A_h=\beta_h+\beta_{N-h} = \frac{1}{N} \sum_{k=0}^{N-1}f_k(e^{-hix_{k}}+e^{-(N-h)ix_k} \]
\[B_h=i(\beta_h-\beta_{N-h}) = \frac{i}{N} \sum_{k=0}^{N-1}f_k(e^{-hix_{k}}-e^{-(N-h)ix_k} \]
\end{proof}$\bigodot$
\begin{theorem}
The constants in the summation
\[S_n(x) = \frac{A_0}{2}+A_n cosnx +\sum_{k=1}^{n-1}(A_kcos(kx)+B_ksin(kx)) \]
Minimise the least square sum.
\[
E(A_0,...,A_N,B_1,...,B_{N-1}) = \sum_{j=0}^{2m-1}(y_j-S_n(x_j))^2 \]

\end{theorem}
\begin{proof}
Similar to the least squares with setting partial derivatives of E wrt $A_k$ and
$B_k$ to zero.ie,
\[
0=\frac{\partial E}{\partial B_k} = 2\sum_{j=0}^{2m-1}(y_j-S_n(x_j))(-sin(kx_j))
\]
so
\begin{eqnarray*}
0&=&\sum_{j=0}^{2m-1}y_j(sinkx_j) - \frac{A_0}{2}\sum_{j=0}^{2m-1}sinkx_j
-A_n\sum_{j=0}^{2m-1}sinkx_jcosnx_j \\
& &- \sum_{l=0}^{2m-1}A_l\sum_{j=0}^{2m-1}sinkx_jcoslx_j 
-\sum_{l=0,l\not=k}^{2m-1}B_l \sum_{j=0}^{2m-1}sinkx_jsinlx_j\\
& & -B_k\sum_{j=0}^{2m-1}(sinkx_j)^2
\end{eqnarray*}
From orthogonality all but the first and last are zero and the previous Theorems
show the final sum is $\frac{N}{2}$
\[
B_k=\frac{1}{m}\sum_{j=0}^{2m-1}y_j(sinkx_j)\] 
\end{proof}$\bigodot$
\textbf{Example:}
See notes
\section{Chebyshev Polynomial}
\begin{definition}
The Chebyshev polynomial $ {T_{n}(x)}$ are orthogonal on $(-1,1)$ w.r.t. the weight function $w(x) = (1-x^2)^{-\frac{1}{2}}$.
\end{definition}
\begin{definition}
$\{ \phi_0,\phi_1,...,\phi_n  \}$ is said to be an orthogonal set of functions,
for the interval [a,b] with respect to the weight function w if
\[
\int^{a}_{b}w(x)\phi_{j}(x)\phi_{k}(x)dx = \left\{\begin{array}{c}0 \ when \ j\not=k\\
\\
\alpha_k > 0 \ when\ j=k \end{array}\right.
\]
If in addition $\alpha=1$ for each k=0,..n the set is said to be orthonormal.
\end{definition}

\subsection*{Defining $T_{n}$}
\[T_0(x) =1, \ \ T_{1}(x)=x \]
\[T_2(x) =2x^2-1, \ \ T_3(x) =4x^3-3x \]  

\[T_{n+1}(x) = 2xT_{n}(x) - T_{n-1}(x) \ \ with \ \ T_{0}=1 \ and \ T_{1}=x \]
$T_{n}(x)$ can also be defined as 
\[T_{n}(x) = cos[n (arccos (x))]\] 
Introducing the substitution $ \theta = (arccos (x))$ changes the equation to
\[T_{n}(\theta(x)) = cosn \theta  \ \ \theta\in[0,\pi] \] 
recurrence relation is derived by noting that

\[T_{n+1}(x) = cos(n\theta) cos\theta -sin(n\theta)sin\theta \] 
and
\[T_{n-1}(x) = cos(n\theta) cos\theta +sin(n\theta)sin\theta \] 
so
\[T_{n+1}(x) = 2cos(n\theta) cos\theta -T_{n-1} \]
Returning to the variable x gives
\[T_{n+1}(x) = 2x T_{n}(x)  -T_{n-1}(x) \]

\[
\int^{1}_{-1}\frac{T_{n}(x)T_{m}(x)}{\sqrt{1-x^2}}dx = \left\{\begin{array}{c}0 \ when \ m\not=n\\
\pi \ n=m=0 \\
\frac{\pi}{2} \ when\ j=k\not=0 \end{array}\right.
\]
\subsection*{Expressing x in terms of $T_{n}$}
\[1=T_{0}, \ \ x = T_1 \]
\[x^2=\frac{1}{2}(T_{0}+T_{2}), \ \ x^2 =\frac{1}{4}(3T_1+T_3) \]
\textbf{Example: see notes}\\
If P(x) is the interpolative poly of degree at most n with nodes at the roots of $T_{n+1}(x)$.
\[
\max_{x\in [-1,1]} |f(x)-P_n(x)| \leq \frac{1}{2^n(n+1)!}\max_{x\in[-1,1]}|f^{n+1}(x)| \ \ \  f \in C^{n+1}[-1,1] \]

\subsection*{Chebyshev Economisation}
\textbf{Purpose:} To approximate a given Nth degree poly
\[
P_{N}(x) = c_0 + c_1 x +...+c_Nx^N \ \ \ c_N\not=0
\]
Approximated on $I+[a,b]$ by a poly $p_{eccon}(x)$ of lower degree.

\begin{enumerate}
\item
Scale $\xi \in [-1,1]$
\[q_{N}(\xi) = P_N(a+\frac{(b-a)}{2}(\xi+1) \]
\item
Get the Chebyshev expansion.
\[q_{N}(\xi) = d_0 + d_1 T_2(\xi) +...+d_NT_{N}(\xi) 
\]
\item
Economise
\[q_{eccon}(\xi) = d_0 + d_1 T_2(\xi) +...+d_kT_{k}(\xi)  \ \ k< N
\]
\item
Expand
\[q_{eccon}(\xi) = e_0 + e_1 \xi +...+e_k\xi^k  
\]
\item
Un-scale $I=[a,b]$
\begin{eqnarray*}
p_{eccon}(\xi) & =&q_{eccon}(-1+2\frac{x-a}{b-a})  \\
&=& \gamma_0 + \gamma_1 x +...+\gamma_kx^k  \ \ a \leq x \leq b
\end{eqnarray*}
\item
Error
\[
|P_{n}(x)-p_{eccon}(x)| \leq |d_{k+1}|+ ...+|d_n| \]
\end{enumerate}
